An introduction to wavelets using R

### 2022-01-11 text/14/wavelets.Rmd
Still more work on wavelets.md

```{r graphing-functions}
graph_with_axis <- function(x, lab, limits=range(x)) {
  par(mar=c(2.6, 0, 0, 0))
  plot((1:2048)/2048, x, 
       xlim=c(0, 1.2), 
       ylim=limits, 
       type="l",
       axes=FALSE)
  axis(side=1, at=(0:4)/4, labels=c(1, 512, 1024, 1536, 2048))
  text(1.1, x[2048], lab, adj=0.5)
}
graph_without_axis <- function(x, lab, limits=range(x)) {
  par(mar=c(0, 0, 0, 0))
  plot((1:2048)/2048, x, 
       xlim=c(0, 1.2), 
       ylim=limits, 
       type="l",
       axes=FALSE)
  text(1.1, x[2048], lab, adj=0.5)
}
```

@@

```{r needle-plot, fig.height=5}
par(mar=c(0,0,0,0))
plot(Re(ft.ecg), Im(ft.ecg), axes=FALSE, type="n")
text(Re(ft.ecg), Im(ft.ecg), 1:2048)
```

@@

```{r orthonormal}
dot.prod <- function(x, y, offset) {
  sum(x*y)
}
dot.prod(g,g)
dot.prod(h,h)
dot.prod(g,h)
```

@@

```{r overlap}
dot.prod(g[1:2],g[3:4])
dot.prod(h[1:2],h[3:4])
dot.prod(g[1:2],h[3:4])
dot.prod(h[1:2],g[3:4])
```

@@

```{r orthogonal-to-linear}
dot.prod(h,1:4)
```

@@

```{r first-level-wavelet}
wt.1 <- mdwt(ecg,g,L=1)
V1 <- wt.1$y[ 1:1024]
W1 <- wt.1$y[1025:2048]
V1[1]
dot.prod(ecg[1:4], g)
V1[2]
dot.prod(ecg[3:6], g)
W1[1]
dot.prod(ecg[1:4], h)
W1[2]
dot.prod(ecg[3:6], h)
wt.2 <- mdwt(ecg,g,L=2)
V2 <- wt.2$y[1:512]
W2 <- wt.2$y[513:1024]
W1_check2 <- wt.2$y[1025:2048]
V2[1]
dot.prod(V1[1:4], g)
V2[2]
dot.prod(V1[3:6], g)
W2[1]
dot.prod(V1[1:4], h)
W2[2]
dot.prod(V1[3:6], h)

level_wavelet <- function(wt.rwt, i) {
  v <-wt.rwt$y
  v[(2^(i-1)):(2^i)] <- 0
  as.vector(midwt(v, d4.rwt$h.0, L=9)$x)
}
```

@@

@@

```{r down-sampling}
dot.prod(ecg, g,1)
V1[1]
dot.prod(ecg, g,3)
V1[2]
dot.prod(ecg, g,5)
V1[3]
```

@@

```{r W1-coefficients}
dot.prod(ecg,-h,1)
W1[1]
dot.prod(ecg,-h,3)
W1[2]
dot.prod(ecg,-h,5)
W1[3]
```

@@

A second pass applies the same pair of filters to V1. This creates 512 coefficients at the second highest frequency (W2) and 512 coefficients containing everything but W1 and W2 (V2).

```{r W2-coefficients}
wt.2 <- mdwt(ecg,g,L=2)
V2 <- wt.2$y[ 1: 512]
W2 <- wt.2$y[513:1024]
dot.prod(V1, g,1)
V2[1]
dot.prod(V1, g,3)
V2[2]
dot.prod(V1, g,5)
V2[3]
```

The V2 coefficients represent the running dot product (with downsampling) of V1 and g.

```{r V2-coefficients}
dot.prod(V1,-h,1)
W2[1]
dot.prod(V1,-h,3)
W2[2]
dot.prod(V1,-h,5)
W2[3]
```

The W2 coefficients represent the running dot product (with downsampling) of V1 and h.

```{r}
all(wt.2$y[1025:2048]==wt.1$y[1025:2048])
```

The second level of the wavelet decomposition preserves all the W1 coefficients in the back half of the wavelet decomposition.

wt.2a <- mdwt(V1,g,L=1)
length(wt.2a$y)
[1] 1024
all(wt.2a$y==wt.2$y[1:1024])
[1] TRUE

You get the exact same wavelet decomposition if you apply a level 1 decomposition on V1.

@@

wt.3 <- mdwt(ecg,g,L=3)
W3 <- wt.3$y[257:512]
V3 <- wt.3$y[ 1:256]

Again for convenience, split apart the decomposition. At this level split into the first eight (1:256) and the second eighth (257:512).

dot.prod(V2, g,1),V3[1],"V3[1]")
V3[1]
By hand -0.28061
Using R -0.28061
dot.prod(V2, g,3),V3[2],"V3[2]")
V3[2]
By hand -0.2404948
Using R -0.2404948
dot.prod(V2, g,5),V3[3],"V3[3]")
V3[3]
By hand -0.2919882
Using R -0.2919882

Are you starting to see the pattern? The V3 coefficients represent the running dot product (with downsampling) of V2 and g.

dot.prod(V2,-h,1),W3[1],"W3[1]")
W3[1]
By hand -0.002777659
Using R -0.002777659
dot.prod(V2,-h,3),W3[2],"W3[2]")
W3[2]
By hand 0.03857375
Using R 0.03857375
dot.prod(V2,-h,5),W2[3],"W3[3]")
W3[3]
By hand 0.0963032950
Using R -0.0003374554

Ditto for the W3 coefficients.

all(wt.3$y[ 513:1024]==wt.2$y[ 513:1024])
[1] TRUE

The W2 coefficients are stored in the same spot for the level 2 and the level 3 decomposition.

wt.3a <- mdwt(V2,g,L=1)
length(wt.3a)
[1] 2
all(wt.3a$y==wt.3$y[1:512])
[1] TRUE

Finally, you get the exact same values by applying a level 1 decomposition to V2.

You keep on going until you get just about run out of V's.

wt.4 <- mdwt(ecg,g,L=4)
W4 <- wt.4$y[129:256]
V4 <- wt.4$y[ 1:128]
wt.5 <- mdwt(ecg,g,L=5)
W5 <- wt.5$y[ 65:128]
V5 <- wt.5$y[ 1: 64]
wt.6 <- mdwt(ecg,g,L=6)
W6 <- wt.6$y[ 33: 64]
V6 <- wt.6$y[ 1: 32]
wt.7 <- mdwt(ecg,g,L=7)
W7 <- wt.7$y[ 17: 32]
V7 <- wt.7$y[ 1: 16]
wt.8 <- mdwt(ecg,g,L=8)
W8 <- wt.8$y[ 9: 16]
V8 <- wt.8$y[ 1: 8]

A level 8 decomposition works pretty well. It has 8 V8's, 8 W8's, 16 W7's, 32 W6's, 64 W5's, 128 W4's, 256 W3's, 512 W2's and 1,024 W1's. The W's store all the information at 8 decreasing frequency levels, and as the frequency decreases from the highest frequency (W1) to lower frequencies, the number of coefficients decreases also. We could take this a couple of steps further, but there is no appreciable benefit to running the wavelet decomoposition all the way to the top, and the edge effects loom very large at the top.

@@

W8d <- dilute(W8)
V8d <- dilute(V8)
 dot.prod(W8d,g,14)+dot.prod(V8d,h,14),V7[1],"V7[1]")
V7[1]
By hand -3.315508
Using R -3.315508
-dot.prod(W8d,g,15)-dot.prod(V8d,h,15),V7[2],"V7[2]")
V7[2]
By hand -2.398146
Using R -2.398146
 dot.prod(W8d,g,16)+dot.prod(V8d,h,16),V7[3],"V7[3]")
V7[3]
By hand -0.2769286
Using R -0.2769286
-dot.prod(W8d,g, 1)-dot.prod(V8d,h, 1),V7[4],"V7[4]")
V7[4]
By hand 1.782748
Using R 1.782748
 dot.prod(W8d,g, 2)+dot.prod(V8d,h, 2),V7[5],"V7[5]")
V7[5]
By hand 5.25007
Using R 5.25007
-dot.prod(W8d,g, 3)-dot.prod(V8d,h, 3),V7[6],"V7[6]")
V7[6]
By hand 4.7496
Using R 4.7496

@@

W7d <- dilute(W7)
V7d <- dilute(V7)
 dot.prod(W7d,g,30)+dot.prod(V7d,h,30),V6[1],"V6[1]")
V6[1]
By hand -2.096562
Using R -2.096562
-dot.prod(W7d,g,31)-dot.prod(V7d,h,31),V6[2],"V6[2]")
V6[2]
By hand -2.294492
Using R -2.294492
 dot.prod(W7d,g,32)+dot.prod(V7d,h,32),V6[3],"V6[3]")
V6[3]
By hand -2.441797
Using R -2.441797
-dot.prod(W7d,g, 1)-dot.prod(V7d,h, 1),V6[4],"V6[4]")
V6[4]
By hand -1.265345
Using R -1.265345
 dot.prod(W7d,g, 2)+dot.prod(V7d,h, 2),V6[5],"V6[5]")
V6[5]
By hand -0.6650157
Using R -0.6650157
-dot.prod(W7d,g, 3)-dot.prod(V7d,h, 3),V6[6],"V6[6]")
V6[6]
By hand 0.08738041
Using R 0.08738041

@@

W1d <- dilute(W1)
V1d <- dilute(V1)
 dot.prod(W1d,g,2046)+dot.prod(V1d,h,2046),ecg[1],"ecg[1]")
ecg[1]
By hand -0.104773
Using R -0.104773
-dot.prod(W1d,g,2047)-dot.prod(V1d,h,2047),ecg[2],"ecg[2]")
ecg[2]
By hand -0.093136
Using R -0.093136
 dot.prod(W1d,g,2048)+dot.prod(V1d,h,2048),ecg[3],"ecg[3]")
ecg[3]
By hand -0.0815
Using R -0.0815
-dot.prod(W1d,g, 1)-dot.prod(V1d,h, 1),ecg[4],"ecg[4]")
ecg[4]
By hand -0.116409
Using R -0.116409
 dot.prod(W1d,g, 2)+dot.prod(V1d,h, 2),ecg[5],"ecg[5]")
ecg[5]
By hand -0.0815
Using R -0.0815
-dot.prod(W1d,g, 3)-dot.prod(V1d,h, 3),ecg[6],"ecg[6]")
ecg[6]
By hand -0.116409
Using R -0.116409

@@

wt.img.1 <- mdwt(img,g,L=1)
sum(img[1:8,1:8]*outer(g,g)),wt.img.1$y[ 1, 1],"V1")
V1
By hand -1.580973
Using R -1.580973
sum(img[1:8,1:8]*outer(h,h)),wt.img.1$y[129,129],"W1")
W1
By hand -0.002825861
Using R -0.002825861
sum(img[1:8,1:8]*outer(h,g)),wt.img.1$y[129, 1],"W1")
W1
By hand -0.01225749
Using R -0.01225749
sum(img[1:8,1:8]*outer(g,h)),wt.img.1$y[ 1,129],"W1")
W1
By hand -0.005042734
Using R -0.005042734

>>>>

```{r graphing-functions}
graph_with_axis <- function(x, lab, limits=range(x)) {
  par(mar=c(2.6, 0, 0, 0))
  plot((1:2048)/2048, x, 
       xlim=c(0, 1.2), 
       ylim=limits, 
       type="l",
       axes=FALSE)
  axis(side=1, at=(0:4)/4, labels=c(1, 512, 1024, 1536, 2048))
  text(1.1, x[2048], lab, adj=0.5)
}
graph_without_axis <- function(x, lab, limits=range(x)) {
  par(mar=c(0, 0, 0, 0))
  plot((1:2048)/2048, x, 
       xlim=c(0, 1.2), 
       ylim=limits, 
       type="l",
       axes=FALSE)
  text(1.1, x[2048], lab, adj=0.5)
}
needle_plot <- function(y, lab, x=seq(0, 1, along.with=y)) {
  par(mar=c(0, 0, 0, 0))
  plot(
    x, y, 
    xlim=c(0, 1.2), 
    type="n",
    axes=FALSE)
  segments(x, 0, x, y)
  text(1.1, 0, lab)
}
```

@@

```{r complex-plot, fig.height=5}
par(mar=c(0,0,0,0))
plot(Re(ft.ecg), Im(ft.ecg), axes=FALSE, type="n")
text(Re(ft.ecg), Im(ft.ecg), 1:2048)
```

@@

```{r orthonormal}
# The inline function %*% calculates a dot product of two vectors.
g %*% g
h %*% h
g %*% h
```

@@

```{r overlap}
g[1:2] %*% g[3:4]
h[1:2] %*% h[3:4]
g[1:2] %*% h[3:4]
h[1:2] %*% g[3:4]
```

@@

```{r orthogonal-to-linear}
h %*% (1:4)
```

@@

@@

Information about the highest frequency events are captured in W1 and information aobut the remainder of the signal are captured in V1. Notice how the largest absolute values in W1 are aligned with some of the very sharp spikes in the signal. The large absolute values in V1 correspond to some of the spikes in the signal that are not quite at the highest frequency as well as low frequency oscillations.

```{r first-level-wavelet}
wt.1 <- mdwt(ecg, g, L=1)
V1 <- wt.1$y[   1:1024]
W1 <- wt.1$y[1025:2048]
{
  needle_plot(V1, "V1")
  needle_plot(W1, "W1")
  graph_without_axis(ecg, "Original\nsignal")
}
```

@@

@@

```{r first-level-dot-products}
c(V1[   1], g %*% ecg[1:4])
c(V1[   2], g %*% ecg[3:6])
c(V1[   3], g %*% ecg[5:8])
c(V1[1023], g %*% ecg[2045:2048])
c(V1[1024], g %*% ecg[c(2047:2048, 1:2)])

c(W1[   1], -h %*% ecg[1:4])
c(W1[   2], -h %*% ecg[3:6])
c(W1[   3], -h %*% ecg[5:8])
c(W1[1023], -h %*% ecg[2045:2048])
c(W1[1024], -h %*% ecg[c(2047:2048, 1:2)])
```

@@

A second pass applies the same pair of filters to V1. This creates 512 coefficients at the second highest frequency (W2) and 512 coefficients containing everything but W1 and W2 (V2). The last 1,024 coefficients are the same highest frequency values (W1) that were computed in the first pass.

```{r second-level-wavelet}
wt.2 <- mdwt(ecg, g, L=2)
V2 <- wt.2$y[1:512]
W2 <- wt.2$y[513:1024]
quality_check_W1 <- wt.2$y[1025:2048]
all(quality_check_W1==W1)
{
  needle_plot(V2, "V2")
  needle_plot(W2, "W2")
  needle_plot(W1, "W1")
  graph_without_axis(ecg, "Original\nsignal")
}
```

```{r second-level-dot-products}
c(V2[  1], g %*% V1[1:4])
c(V2[  2], g %*% V1[3:6])
c(V2[  3], g %*% V1[5:8])
c(V2[511], g %*% V1[1021:1024])
c(V2[512], g %*% V1[c(1023:1024, 1:2)])

c(W2[  1], -h %*% V1[1:4])
c(W2[  2], -h %*% V1[3:6])
c(W2[  3], -h %*% V1[5:8])
c(W2[511], -h %*% V1[1021:1024])
c(W2[512], -h %*% V1[c(1023:1024, 1:2)])
```

You could get the same results as a level 2 wavelet transform by applying a level 1 wavelet transform to V1.

```{r transform-V1}
quality_check <- mdwt(V1, g, L=1)
quality_check_V2 <- quality_check$y[  1: 512]
quality_check_W2 <- quality_check$y[513:1024]
all(quality_check_V2==V2)
all(quality_check_W2==W2)
```

@@

A third pass applies the same pair of filters to V2. This creates 256 coefficients at the third highest frequency (W3) and 256 coefficients containing information about all the lower frequency events (V3). These 512 values are followed by the 512 W2 coefficients and the 1,024 W3 coefficients.

```{r third-level-wavelet}
wt.3 <- mdwt(ecg, g, L=3)
V3 <- wt.3$y[1:256]
W3 <- wt.3$y[257:512]
W2_check3 <- wt.3$y[513:1024]
W1_check3 <- wt.3$y[1025:2048]
all(W2_check3==W2)
all(W1_check3==W1)
{
  needle_plot(V3, "V3")
  needle_plot(W3, "W3")
  needle_plot(W2, "W2")
  needle_plot(W1, "W1")
  graph_without_axis(ecg, "Original\nsignal")
}
```

Again, large coefficients in W3 are associated with more of the spikes in the original signal--spikes that are not quite as abrupt, perhaps, as the spikes captured in W1 and W2.

The values of W3 and V3 are computed as before, but using V2 instead of V1 or the original signal.

```{r third-level-dot-products}
c(V3[  1], g %*% V2[1:4])
c(V3[  2], g %*% V2[3:6])
c(V3[  3], g %*% V2[5:8])
c(V3[255], g %*% V2[509:512])
c(V3[256], g %*% V2[c(511:512, 1:2)])

c(W3[  1], -h %*% V2[1:4])
c(W3[  2], -h %*% V2[3:6])
c(W3[  3], -h %*% V2[5:8])
c(W3[255], -h %*% V2[509:512])
c(W3[256], -h %*% V2[c(511:512, 1:2)])
```

Here's an outline of how the wavelet transform works, level by level. Take the original 2,048 signal values and transform them into 1.024 very high frequency values (W1) and what's left over (V1). Take the 1,024 V1 values and split them into 512 high frequency values (W2) and what's left over (V2). Take those V2 values and split them. Keep on spliting until there are only a handful of V values left. I won't show the details of further calculation, but here is what the graph looks like.

```{r ninth-level-wavelet}
wt.9 <- mdwt(ecg, g, L=9)
V9 <- wt.9$y[   1:   4]
W9 <- wt.9$y[   5:   8]
W8 <- wt.9$y[   9:  16]
W7 <- wt.9$y[  17:  32]
W6 <- wt.9$y[  33:  64]
W5 <- wt.9$y[  65: 128]
W4 <- wt.9$y[ 129: 256]
W3 <- wt.9$y[ 257: 512]
W2 <- wt.9$y[ 513:1024]
W1 <- wt.9$y[1025:2048]
{
  needle_plot(V9, "V9")
  needle_plot(W9, "W9")
  needle_plot(W8, "W8")
  needle_plot(W7, "W7")
  needle_plot(W6, "W6")
  needle_plot(W5, "W5")
  needle_plot(W4, "W4")
  needle_plot(W3, "W3")
  needle_plot(W2, "W2")
  needle_plot(W1, "W1")
  graph_without_axis(ecg, "Original\nsignal")
}
```

```{r}
level_wavelet <- function(wt.rwt, i) {
  v <-wt.rwt$y
  v[(2^(i-1)):(2^i)] <- 0
  as.vector(midwt(v, d4.rwt$h.0, L=9)$x)
}
```

### Reconstruction of the original signal

@@

First you dilute the W and V values at the top of the pyramid.

```{r reconstuct V8}
V9_diluted <- c(V9[1], 0, V9[2], 0, V9[3], 0, V9[4], 0)
W9_diluted <- c(W9[1], 0, W9[2], 0, W9[3], 0, W9[4], 0)

wt.8 <- mdwt(ecg, g, L=8)
V8 <- wt.8$y[1:8]

c(V8[1], rev(h) %*% V9_diluted[1:4] + rev(g) %*% W9_diluted[1:4])
c(V8[2], rev(h) %*% V9_diluted[2:5] + rev(g) %*% W9_diluted[2:5])
```

W8d <- dilute(W8)
V8d <- dilute(V8)
 W8d, g, 14)+V8d,h,14),V7[1],"V7[1]")
V7[1]
By hand -3.315508
Using R -3.315508
-W8d, g, 15)-V8d,h,15),V7[2],"V7[2]")
V7[2]
By hand -2.398146
Using R -2.398146
 W8d, g, 16)+V8d,h,16),V7[3],"V7[3]")
V7[3]
By hand -0.2769286
Using R -0.2769286
-W8d, g,  1)-V8d,h, 1),V7[4],"V7[4]")
V7[4]
By hand 1.782748
Using R 1.782748
 W8d, g,  2)+V8d,h, 2),V7[5],"V7[5]")
V7[5]
By hand 5.25007
Using R 5.25007
-W8d, g,  3)-V8d,h, 3),V7[6],"V7[6]")
V7[6]
By hand 4.7496
Using R 4.7496

@@

W7d <- dilute(W7)
V7d <- dilute(V7)
 W7d, g, 30)+V7d,h,30),V6[1],"V6[1]")
V6[1]
By hand -2.096562
Using R -2.096562
-W7d, g, 31)-V7d,h,31),V6[2],"V6[2]")
V6[2]
By hand -2.294492
Using R -2.294492
 W7d, g, 32)+V7d,h,32),V6[3],"V6[3]")
V6[3]
By hand -2.441797
Using R -2.441797
-W7d, g,  1)-V7d,h, 1),V6[4],"V6[4]")
V6[4]
By hand -1.265345
Using R -1.265345
 W7d, g,  2)+V7d,h, 2),V6[5],"V6[5]")
V6[5]
By hand -0.6650157
Using R -0.6650157
-W7d, g,  3)-V7d,h, 3),V6[6],"V6[6]")
V6[6]
By hand 0.08738041
Using R 0.08738041

@@

W1d <- dilute(W1)
V1d <- dilute(V1)
 W1d, g, 2046)+V1d,h,2046),ecg[1],"ecg[1]")
ecg[1]
By hand -0.104773
Using R -0.104773
-W1d, g, 2047)-V1d,h,2047),ecg[2],"ecg[2]")
ecg[2]
By hand -0.093136
Using R -0.093136
 W1d, g, 2048)+V1d,h,2048),ecg[3],"ecg[3]")
ecg[3]
By hand -0.0815
Using R -0.0815
-W1d, g,  1)-V1d,h, 1),ecg[4],"ecg[4]")
ecg[4]
By hand -0.116409
Using R -0.116409
 W1d, g,  2)+V1d,h, 2),ecg[5],"ecg[5]")
ecg[5]
By hand -0.0815
Using R -0.0815
-W1d, g,  3)-V1d,h, 3),ecg[6],"ecg[6]")
ecg[6]
By hand -0.116409
Using R -0.116409

@@

wt.img.1 <- mdwt(img, g, L=1)
sum(img[1:8,1:8]*outer(g,g)),wt.img.1$y[ 1, 1],"V1")
V1
By hand -1.580973
Using R -1.580973
sum(img[1:8,1:8]*outer(h,h)),wt.img.1$y[129,129],"W1")
W1
By hand -0.002825861
Using R -0.002825861
sum(img[1:8,1:8]*outer(h,g)),wt.img.1$y[129, 1],"W1")
W1
By hand -0.01225749
Using R -0.01225749
sum(img[1:8,1:8]*outer(g,h)),wt.img.1$y[ 1,129],"W1")
W1
By hand -0.005042734
Using R -0.005042734
